---
title: "WASH in Schools in  Piracema, Brazil"
author: "Kasandra Poague"
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
editor_options: 
  chunk_output_type: console
editor: visual
---

## Brazilian National School Census

Since 2014 INEP (*Instituto National de Estudos e Pesquisas Educacionais Anísio Teixeira)*, the Brazilian government entity responsible for the surveillance of schools, has been collecting and releasing information on an annual basis on the infrastructure of all basic education institutions in the country, regardless of the level of education, locality (rural vs urban) and management model (private or public) (INEP, 2021). Every year, the school's principals, headteachers, or the person in charge must reply to a self-reported survey sent by the INEP. The questionnaire contains 62 questions and should be filled out between June and August, considering the last Wednesday of May as the reference date for data collection (INEP, 2021). The final results are released at the end of January and the beginning of February from the following year. For the year 2020, due to the extraordinary situation of the COVID-19 pandemic, the reference date for collecting the data was postponed to March 11 2020, one day before the national closure of schools was established. Hence, the 2020 BNSC dataset describes the state of schools' infrastructure in Brazil right before they were closed. For 2021 and 2022, the reference date for data collection returned to the original standard (last Wednesday of May of 2021 and 2022). Thus, the 2021 BNSC, which were released on February 18 2022 reflect the impacts of the pandemic and the school closure on the state of schools' infrastructure in Brazil, while the 2022 BNSC released on and February 08 2023 indicates the situation of schools at the end of the pandemic.

We adopted the word "pre" to designate the state of schools before the beginning of the pandemic  (2020 BNSC), "peri" to refer to the state of schools during the pandemic (2021 BNSC) and "post" to indicate the conditions of schools moving forward from the COVID-19 pandemic (at the end of the pandemic).

## Piracema

Located in the southeast region of Brazil, in the state of Minas Gerais, the municipality of Piracema is a small size city adding up to 6,700 inhabitants (IBGE, 2023). It ranks as the 3,734º and 492º smallest municipality among all Brazilian territories (out of 5,570 municipalities) and in Minas Gerais (out of 853 municipalities), respectively (IBGE, 2023). Piracema is located approximately 120 km away from the capital of the state its state (Belo Horizonte), and it is inaccessible by public transportation (IBGE, 2023). Piracema will be the study area for the next phase of the research (collection of primary data).

## Goals

The main goal of this study was to describe the profile of schools from the basic education system in the municipality of Piracema (Minas Gerais, Brazil). Moreover, we also aimed to compare the characteristics of schools, with a special concern to WASH, pre- (2020), peri- (2021) and post-COVID-19 pandemic (2022) to evaluate to what extend schools in Piracema made progress in providing WASH since the beginning of the COVID-19 pandemic.

This study is a sub-project of a PhD project and an initial study comparing the BNSC from 2020 and 2021 has already been conducted and published by the author of this project (for more details see references). Findings will be essential for the next phase of the research, which will be the collection of primary data in schools in the municipality of Piracema through qualitative methods (interviews, on-spot observations and art-based research).

## References and sources

IBGE, 2023. Piracema. <https://cidades.ibge.gov.br/brasil/mg/piracema/panorama>

INEP, 2021. Censo da Educação Básica 2020 - Resumo Técnico. https://download.inep.gov.br/publicacoes/institucionais/estatisticas_e_indicadores/resumo_tecnico_censo_escolar_2020.pdf81.

INEP, 2022. Censo Escolar. Microdados do Censo Escolar de Educação Básica.  2021  <https://www.gov.br/inep/pt-br/acesso-a-informacao/dados-abertos/microdados/censo-escolar>.

**Poague, K. I. H. M**.; Blanford, J.I.; Martinez, J.A.; Anthonj, C. Water, sanitation and hygiene (WASH) in schools in Brazil pre-and peri-COVID-19 pandemic: are schools making any progress? Int J Hyg Environ Health, 247, 114069, <https://doi.org/10.1016/j.ijheh.2022.114069>, 2023.

## Import

```{r}
 library(readr) 
 library(tidyverse) 
 library(janitor)
 
Piracema_2020<-read_csv(here::here("data/raw/Piracema_2020_clean.csv"))
Piracema_2021<-read_csv(here::here("data/raw/Piracema_2021_clean.csv"))
Piracema_2022<-read_csv(here::here("data/raw/Piracema_2022_clean.csv"))
```

## Data Anonymation

```{r}
Piracema_2020<-Piracema_2020[,-c(2,3,4,5,6,9)]
Piracema_2021<-Piracema_2021[,-c(2,3,4,5,6,9)]
Piracema_2022<-Piracema_2022[,-c(2,3,4,5,6,9)]
```

## Combining datasets

```{r}
Piracema_2020_2022<-full_join(Piracema_2020,Piracema_2021) |> full_join(Piracema_2022)
```

## Changing variable type

All variables so far are numeric.

```{r}
str(Piracema_2020_2022)
```

Changing the variables to factors and characters.

```{r}

Piracema_2020_2022_character<- data.frame(lapply(Piracema_2020_2022 [1:3], as.character)) 
str(Piracema_2020_2022_character)

Piracema_2020_2022_factor<- data.frame(lapply(Piracema_2020_2022 [4:6], as.factor)) 
str(Piracema_2020_2022_factor)

```

Calculating the percentage of students to replace the number of students

```{r}
Piracema_2020_2022$PC_MENINA<- Piracema_2020_2022$QT_MAT_BAS_FEM/Piracema_2020_2022$QT_MAT_BAS*100

Piracema_2020_2022$PC_BRANCO<- Piracema_2020_2022$QT_MAT_BAS_BRANCA/Piracema_2020_2022$QT_MAT_BAS*100

Piracema_2020_2022$PC_PARDO<- Piracema_2020_2022$QT_MAT_BAS_PARDA/Piracema_2020_2022$QT_MAT_BAS*100

Piracema_2020_2022$PC_PRETO<- Piracema_2020_2022$QT_MAT_BAS_PRETA/Piracema_2020_2022$QT_MAT_BAS*100

Piracema_2020_2022$PC_INDIO<- Piracema_2020_2022$QT_MAT_BAS_INDIGENA/Piracema_2020_2022$QT_MAT_BAS*100

Piracema_2020_2022$PC_ASIATICO<- Piracema_2020_2022$QT_MAT_BAS_AMARELA/Piracema_2020_2022$QT_MAT_BAS*100


Piracema_2020_2022_double<- data.frame(lapply(Piracema_2020_2022 [40:45], as.double)) 

str(Piracema_2020_2022_double)
```

Now getting all together.

```{r}
Piracema_2020_2022_types<-cbind(Piracema_2020_2022_character,Piracema_2020_2022_factor,Piracema_2020_2022_double)|> 
  cbind(Piracema_2020_2022[7:25]) |> 
  cbind(Piracema_2020_2022[34:39])

str(Piracema_2020_2022_types)
```

## Checking for changes across the years

Now that we have our data ready, lets analyze if there were any changing during those three years.

@tbl-drinking-water shows the percentage of schools with drinking water and their water source throughout the years.

```{r}
#| label: tbl-drinking-water
#| tbl-cap: "Schools with drinking water and type of water source (%)"
tbl_drinking_water<- Piracema_2020_2022_types |> 
  group_by(NU_ANO_CENSO) |>
  summarise( 
   Drinking_water = sum(IN_AGUA_POTAVEL)/7*100,
   Public_newtork = sum(IN_AGUA_REDE_PUBLICA)/7*100,
   Borehole = sum(IN_AGUA_POCO_ARTESIANO)/7*100,
   Cacimba = sum(IN_AGUA_CACIMBA)/7*100,
   Superficial_source = sum(IN_AGUA_FONTE_RIO)/7*100,
   No_source = sum(IN_AGUA_INEXISTENTE)/7*100
  ) 

tbl_drinking_water
```

@tbl-bathroom presents the percentage of schools equipped with sanitary facilities.

```{r}
#| label: tbl-bathroom
#| tbl-cap: "Schools with sanitation facilities (%)"
tbl_bathroom<- Piracema_2020_2022_types |> 
  group_by(NU_ANO_CENSO) |>
  summarise( 
   Bathroom = sum(IN_BANHEIRO)/7*100,
   Bathroom_Youngest_children = sum(IN_BANHEIRO_EI)/7*100,
   Bathroom_Disable = sum(IN_BANHEIRO_PNE)/7*100,
   Bathroom_with_showers = sum(IN_BANHEIRO_CHUVEIRO)/7*100,
   Bathroom_for_staff = sum(IN_BANHEIRO_FUNCIONARIOS)/7*100,
  )

tbl_bathroom
```

While @tbl-sanitation indicates the type of sanitation solution.

```{r}
#| label: tbl-sanitation
#| tbl-cap: "Schools by type of sanitation solution (%)"
tbl_sanitation<- Piracema_2020_2022_types |> 
  group_by(NU_ANO_CENSO) |>
  summarise( 
   Sewerage = sum(IN_ESGOTO_REDE_PUBLICA)/7*100,
   Septic_Tank= sum(IN_ESGOTO_FOSSA_SEPTICA)/7*100,
   No_sanitation = sum(IN_ESGOTO_INEXISTENTE)/7*100,
   Improved_sanitation = sum(IN_ESGOTO_REDE_PUBLICA+IN_ESGOTO_FOSSA_SEPTICA)/7*100,
  )

tbl_sanitation
```

And last, what the schools do with their waste (@tbl-waste).

```{r}
#| label: tbl-waste
#| tbl-cap: "Schools by type of waste management (%)"
tbl_waste<- Piracema_2020_2022_types |> 
  group_by(NU_ANO_CENSO) |>
  summarise( 
   Collection = sum(IN_LIXO_SERVICO_COLETA)/7*100,
   Public_disposal= sum(IN_LIXO_DESTINO_FINAL_PUBLICO)/7*100,
   Burned = sum(IN_LIXO_QUEIMA)/7*100,
   Buried = sum(IN_LIXO_ENTERRA)/7*100,
   Disposed_another_area = sum(IN_LIXO_DESCARTA_OUTRA_AREA)/7*100,
  )

tbl_waste
```

We observed that between 2020 and 2022 the only changes were only changes in the frequency of schools with bathroom for people with disabilities or limited mobility (from 71.4% to 85.7%) and bathrooms exclusively for staff (from 57.1% to 85.7%). So now, lets find all which schools were those and if by any chance they are the same institution.

```{r}
which(Piracema_2020_2022_types$IN_BANHEIRO_PNE == 1)

```

**Answer:** school in position(row) 6.

```{r}
which(Piracema_2020_2022_types$IN_BANHEIRO_FUNCIONARIOS == 1)

```

**Answer:** schools in positions(rows) 3 and 6.
